An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging

### 2014-04-24 HowtoHowtoCapon/submission/final/article.tex

\newif\ifPeerReview\PeerReviewfalse             % Whether to create the PeerReview version or
                                                % Journal version
\newif\ifFlatArchive\FlatArchivefalse           % Whether archive is flat (messy) or contain 
                                                % subfolders for graphics etc.
\newif\ifFloatAtEnd\FloatAtEndfalse             % Available in PeerReview mode:

\graphicspath{{./Paper1/}}

\definecolor{tabBlue}{HTML}{AACCFF}

@@

@@

@@

%Math macros:
\newcommand\diff[2]{\frac{\kw{d}\;\textstyle #1\scriptstyle}{\kw{d\parm{#2}}}\displaystyle}
\newcommand\ddiff[2]{\frac{\kw{d^2}\;\displaystyle #1\scriptstyle}{\kw{d\parm{#2}}^2}\displaystyle}

\renewcommand\d[1]{\scriptstyle\kw{\;d\parm{#1}}\displaystyle}

% These commands are mutually exclusive. Remember to "renew" in v2.
\newcommand\intb[4]{\int\limits_{#3}^{#4} #1 \d{#2}} % \int{exp}{var}{from}{to}
\newcommand\intl[3]{\int\limits_{#3} #1 \d{#2}} % \int{exp}{var}{for all}
\newcommand\intu[2]{\int #1 \d{#2}} % \int{exp}{var}{for all}

\newcommand\T{^{\scriptscriptstyle T}}
\renewcommand\H{^{\scriptscriptstyle H}}

\renewcommand\vec[1]{\boldsymbol{#1}}
\renewcommand\mat[1]{\boldsymbol{#1}}

@@

\newcommand*\A{\mat A}
\newcommand*\B{\mat B}
\newcommand*\C{\mat C}
\newcommand*\E{\mat E}
% \renewcommand*\H{\mat H}
\renewcommand*\P{\mat P}
\newcommand*\eP{\mat{\hat P}}
\renewcommand*\R{\mat R}
\newcommand*\Ri{\R^{-1}}
\newcommand*\eR{\mat{\hat R}}
\newcommand*\eRi{\hat{\mat R}\;\!^{-1}}
\newcommand*\Navg{N_\text{avg}}
\newcommand*\W{\mat W}
\newcommand*\X{\mat X}
\newcommand*\Xd{\X_{\!\Delta}}
\newcommand*\Y{\mat Y}

@@

\newcommand\D{\vec\nabla} %Del: Vector differential operator - nabla
\newcommand\Dx{\vec\nabla\times}
\newcommand\Dd{\vec\nabla\cdot}

@@

\begin{abstract}
% Two limiting factors of modern phased array sonar imaging systems are the angular resolution and contrast. Adaptive beamformers target these limitatitons by weighting sensors dynamically to fit the impinging wavefield.

@@

@@

Data driven methods have been introduced in various imaging fields over the years in attempts to improve image quality. One such method is the Minimum Variance Distortionless Response (MVDR) beamformer. When compared to conventional methods, MVDR is often able to improve image contrast and resolution, as shown in e.g. radar~\cite{Benitz1997}, ultrasound~\cite{Synnevag2007}, and active sonar~\cite{Blomberg2012a,Blomberg2011,Dursun2009,Lo2004}.

@@

The implementation is adapted to a well sampled 100\;kHz wideband active sonar system operating in the near field. The system has a range resolution comparable to the sample period so our MVDR implementation runs near single snapshot mode. To deal with this and avoid signal cancellation we compute and average the covariance estimates from overlapping subarrays~\cite{Kailath1985}. This may resemble techniques such as "subarray MVDR"~\cite{Chapman1976}, but unlike that technique we are not using subarrays to reduce complexity but to make MVDR in an active system possible. In related work, we investigated using the same MVDR GPU implementation in cardiac ultrasound imaging~\cite{Asen2012,Asen2013}, and in active sonar imaging we have compared the performance of this implementation to comparable adaptive methods~\cite{Buskenes2013}.

@@

%  Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of around 2 orders of magnitude compared to an optimized single-thread implementation in C.

% These devices have in recent years emerged as a flexible and highly potent parallel technology well suited for many DSP tasks, and are available off-the-shelf at reasonable prices\todo{skip this?}.

% These devices have in recent years emerged as a flexible and highly potent parallel technology well suited for many DSP tasks, and are available off-the-shelf at reasonable prices.

@@

% - Adaptive beamformers offers image quality edge, but are usually computationally complex.
% - Who's tried before?
% - We: Active sonar MVDR, GPU, high performance.

@@

\ifPeerReview
\begin{figure}[H]\centering
\includegraphics[width=0.8\linewidth]{gfx/buske1.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=0.8\linewidth]{gfx/beamforming.eps}
\fi%
\caption{Beamforming principle. Signal signature is first removed by matched filtering. Then - prior to summation - a suitable set of delays, $\Delta$, and weights, $w$, are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{beamforming}
\end{figure}
Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (Fig. \ref{beamforming}). Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$. To simplify notation, we make the dependence on $\theta$ implicit from now on.

@@

Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield. In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}. This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\} \in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since we pre-steer our data to every pixel in the image we will always look to broadside, but (\ref{weights}) can be easily generalized by substituting $\1$ with a suitable steering vector. Since $\R$ is unknown, we estimate it by computing a sample covariance matrix $\eR$. In this computation we perform some degree of:
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985},
\item \emph{temporal averaging} over an interval comparable to the pulse length (1-5 samples) to maintain true speckle statistics~\cite{Synnevag2009a}, and
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
\end{itemize}%
%
% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009a}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}. 
%
Combined, these steps will also ensure a numerically well conditioned $\eR$.

To form $\eR$ we will first compute an intermediate sample covariance matrix $\breve{\R}$, which we form using temporal and spatial averaging. To do this we need to segment our array into subarrays. If we let $x_l[n]$ represent the data vector from subarray $l$,
\begin{gather}
\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T,
\end{gather}
then $\breve{\R}$ can be calculated as:
\begin{gather}
\breve{\R}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n']\x_l\H[n'] \in\mathbb{C}^{L,L},\label{spatialR}
\end{gather}
where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.

@@

Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{z}). This weight set is applied to all the subarrays, prior to computing the beamformer output as in (\ref{z}). Or, equivalently, we may apply the weight set to the sum of all the subarrays:
\begin{align}
z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{finalZ}
\end{align}
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske2.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/implementation.eps}
\fi%
\caption{MVDR beamforming. For each pixel in range and azimuth,\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
3. and the beamformer output $z$ is computed from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
\end{figure}
% \begin{figure}[!t]\centering
% \includegraphics[width=\linewidth]{gfx/algorithm_structure.eps}
% \caption{MVDR beamforming. First a spatial covariance matrix is estimated from the delayed data (\ref{spatialR}-\ref{finalR}), then the weights are computed (\ref{weights}) and finally applied to the delayed channel data (\ref{z}).}
% \label{implementation}
% \end{figure}
As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
\begin{enumerate}
\item computing the sample covariance matrix $\eR$ in (\ref{finalR}),
\item computing $\eRi\1$ in (\ref{weights}), and
\item computing the beamformer output $z$ in (\ref{finalZ}).
\end{enumerate}
% This is illustrated in .
% \begin{figure}[!t]\centering
% \includegraphics[width=\linewidth]{gfx/buildR-complex_data-average.eps}
% \caption{.}\label{complexity}
% \end{figure}
Next we will evaluate these steps in terms of arithmetic complexity, and then discuss their mappability to parallel hardware.

@@

If we neglect spatial and temporal averaging, then the computation of $\eR$ is reduced to a single outer product with complexity of O($M^2$), and the inversion is then of O($M^3$). This might lead us to believe that the inversion step is by far the most complex. But if we implement spatial and temporal averaging as in (\ref{spatialR}), then computing $\eR$ is of O($N_K N_L L^2$) and the inversion of O($L^3$). Computing $\eR$ is now the most complex operation whenever $N_LN_K>L$.
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske3.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/mvdr_complexity.eps}
\fi%
\caption{Per-pixel computational complexity of the steps in MVDR beamforming (prior to any optimizations). To avoid signal cancellation in an active sonar system we usually set $L<\frac{M}{2}$, in which region the computation of $\eR$ dominates in terms of arithmetic complexity, especially when performing temporal averaging.}\label{mvdr_complexity}
\end{figure}
To visualize these relations, and include the effects of using complex numbers, we set up complexity formulas that account for the total number of arithmetic operations for each step in the MVDR process (see appendix \ref{mvdr_formulas}). We only excluded diagonal loading performed in the computation of $\eR$, and partial pivoting used in the inversion step, since these operations contribute marginally to the end result. The entire range of possible subarray sizes from $L\in[1,M]$ was finally evaluated, with temporal averaging set to $K\in\{0,1,2\}$, and the number of channels set to $M\in\{8,16,32\}$. The results are shown in \Fig{mvdr_complexity}.

We note how the computation of $\eR$ completely dominates at smaller subarray sizes, that solving $\eRi\1$ only plays a notable role for larger array and subarray sizes, and that the computation of $z$ has a negligible impact on computation time. Also notice how temporal averaging comes with a high computational penalty. This is because building $\eR$ is heavy on complex multiplications, which require 3 times as many arithmetic instructions as a complex addition, and a lot of these are repeated unnecessarily.

% Not corrected for fused multiply accumulate

% Also, the need for partial pivoting has been ignored, as we ensure a numerically well conditioned $\eR$ by applying diagonal loading. Since this is the most common scenario, we will have a thorough look at how it can be optimized.

% and does not take into that does not  feasibility parallel decomposition  memory consumption.

@@

% \begin{itemize}
% \item Keep delay step on CPU, as this is a common step in all beamforming
% \item Increasing matrix size reduces the number of threads per block, since the entire matrix must be in shared memory. 1 matrix 72x72 complex (8 bytes) max per block - 72 threads. same as 207 5x5 matrixes, 5*207=1036 threads. max threads in block 512.
% \end{itemize}

% An important feature of MVDR beamforming, and with beamforming in general, is that each pixel can be computed independently and in an identical fashion. Considering that sonar images often end up having millions of pixels, this makes a compelling case for investigating technologies that support massive data level parallelism. The technology that does this best is perhaps Graphics Processing Units (GPUs). These chips contain up to thousands of small computation cores, which combined deliver superior peak floating point performance compared multi-core systems such as CPUs.

@@

% What about the GPU implementation? This one was tricky. Balance load, avoid inter-thread dependencies when performing iterative summations, keep thread memory consumption at a minimum, and promote collective access patterns (coalesced reads/writes) to global memory.

>>>>

%%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% CONDITIONALS %
%%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\newif\ifPeerReview\PeerReviewtrue              % Whether to create the PeerReview version or
                                                % Journal version
\newif\ifOnlineColor\OnlineColortrue            % Compile online color version?

\newif\ifFlatArchive\FlatArchivefalse           % Whether archive is flat (messy) or contain 
                                                % subfolders for graphics etc.
\newif\ifFloatAtEnd\FloatAtEndfalse             % Available in PeerReview mode:
                                                % Place floats at end of document?
\newif\ifTODO\TODOtrue                        % Use todo notes?

\ifOnlineColor
   \newcommand\figPostfix{_online}              % Colored images online
\else
   \newcommand\figPostfix{_bw}                  % Black and white in journal
\fi

%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% IEEEtran %
%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\ifPeerReview
\documentclass[12pt,journal,draftclsnofoot,onecolumn]{IEEEtran}
% \newcommand\CLASSINPUTbaselinestretch{1.66}     % http://theoval.cmp.uea.ac.uk/~nlct/latex/thesis/node17.html
\else
\documentclass[journal]{IEEEtran}
\fi

% \RequirePackage[latin1]{inputenc}%              % Set input encoding (optionally latin1)
% \RequirePackage[T1]{fontenc}%                 % Set font encoding
% \usepackage[norsk]{babel}

% Single column review mode:
%\documentclass[12pt,journal,onecolumn]{IEEEtran}
%\newcommand\CLASSINPUTbaselinestretch{2}
%\RequirePackage{calc}
%\RequirePackage{fp} 
% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
% Adding the 2cm gutter width to the odd/even side margins:
%\setlength\hoffset{0pt}
%\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
%\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
%\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% IEEE ''APPROVED'' PACKAGES %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\usepackage{cite}

\ifCLASSINFOpdf
   \usepackage[dvips]{graphicx}                 % Might not work. Use 'latex' instead of 
   \ifFlatArchive\else                          % 'pdflatex'
      \graphicspath{./gfx/}
   \fi
\else
   \usepackage[dvips]{graphicx}
   \ifFlatArchive\else
      \graphicspath{./gfx/}
   \fi
\fi

\RequirePackage[table,dvipsnames,svgnames]{xcolor}

\usepackage[cmex10]{amsmath}                    % cmex10 option to be IEEE explore compliant
\interdisplaylinepenalty=2500                   % Allows multiline equations to be broken

% \RequirePackage{amssymb}

\RequirePackage{array}

\ifCLASSOPTIONcompsoc
   \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
\else
   \usepackage[caption=false,font=footnotesize]{subfig}
\fi
\ifCLASSOPTIONcaptionsoff                       % IEEE promoted hack to turn off captions from the 
   \let\MYorigsubfloat\subfloat                 % subfloat package should the captionsoff option
   \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}} % be specified.
\fi

\ifFloatAtEnd
\ifCLASSOPTIONcaptionsoff                       % Places float at the end of the document when the
  \usepackage[nomarkers]{endfloat}              % captionsoff options is specified to IEEEtrans.cls
  \let\MYoriglatexcaption\caption               % (PeerReview mode)
  \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
\fi
\fi

\usepackage{fixltx2e}                           % Fix some twocolumn float problems

%\usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
                                                % (double column figures on top/bottom)

\usepackage{url}                                % Support for handling and breaking URLs

% NOTE: PDF hyperlink and bookmark features are not required in IEEE
%       papers and their use requires extra complexity and work.
\newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
pdftitle={An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging},
pdfsubject={},
pdfauthor={Jo Inge Buskenes},
pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
\ifCLASSINFOpdf
   \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
\else
   \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
   \usepackage{breakurl}                        % Allows 'dvips' driver to break links
\fi

%%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% ADDITIONAL PACKAGES %       
%%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% \ifPeerReview
% \let\OldIncludegraphics{\includegraphics}
% \usepackage{letltxmacro}
% \LetLtxMacro{\OldIncludegrsaphics}{\includegraphics}
% \renewcommand{\includegraphics}[2][]{\OldIncludegraphics[width=\linewidth, #1]{#2}}
% \fi

% % \makeatletter
% % \def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
% % \else\Gin@nat@width\fi}
% % \makeatother
% % \let\Oldincludegraphics\includegraphics
% % \renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}

\usepackage[maxfloats=40]{morefloats}
\newcounter{todoidx}
% \setcounter{todoidx}

\ifTODO
   \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
   \setlength\marginparsep{1pt}
   \setlength\marginparwidth{35pt}
   \newlength\marginparwidthsmall
   \setlength\marginparwidthsmall{\marginparwidth}
   \addtolength\marginparwidthsmall{-7pt}
   \newcommand\todo[1]{%
      \addtocounter{todoidx}{1}%
      {\color{Red}\bf(\thetodoidx{})}%%\fbox{\bf\thetodoidx{}}}%
      \marginpar{%
         {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\scriptsize #1}}}}

   \newcommand\todopar[1]{\fcolorbox{red}{white}{\parbox{0.97\linewidth}{#1}}}
\else
%    \usepackage[disable]{./todonotes} 
   \newcommand\todo[1]{}
\fi

\newenvironment{narrow}[2]{%
\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{#1}%
\setlength{\rightmargin}{#2}%
\setlength{\listparindent}{\parindent}%
\setlength{\itemindent}{\parindent}%
\setlength{\parsep}{\parskip}}%
\item[]}{\end{list}}

\usepackage{float}

\ifOnlineColor
   \definecolor{tabBlue}{HTML}{AACCFF}
\else
   \definecolor{tabBlue}{HTML}{CCCCCC}
\fi

@@

% \DeclareMathOperator*{\argmin}{\text{arg}\;\text{min}}

@@

\newcommand\Grey[1]{{\color{Grey}#1}}
\newcommand\Red[1]{{\color{Red}#1}}
\newcommand\Blue[1]{{\color{Blue}#1}}
\newcommand\DarkBlue[1]{{\color{DarkBlue}#1}}
\newcommand\LightBlue[1]{{\color{LightBlue}#1}}
\newcommand\Brown[1]{{\color{Brown}#1}}
\newcommand\Green[1]{{\color{Green}#1}}
\newcommand\SeaGreen[1]{{\color{SeaGreen}#1}}
\newcommand\Yellow[1]{{\color{yellow}#1}}
\newcommand\Orange[1]{{\color{orange}#1}}

@@

%Math macros:
\newcommand\T{^{\scriptscriptstyle T}}
\renewcommand\H{^{\scriptscriptstyle H}}

\renewcommand\vec[1]{\boldsymbol{#1}}
\newcommand\mat[1]{\boldsymbol{#1}}

@@

\newcommand*\A{\mat A}
\newcommand*\B{\mat B}
\newcommand*\C{\mat C}
\newcommand*\E{\mat E}
\renewcommand*\P{\mat P}
\newcommand*\eP{\mat{\hat P}}
\newcommand*\R{\mat R}
\newcommand*\Ri{\R^{-1}}
\newcommand*\eR{\mat{\hat R}}
\newcommand*\eRi{\hat{\mat R}\;\!^{-1}}
\newcommand*\Navg{N_\text{avg}}
\newcommand*\W{\mat W}
\newcommand*\X{\mat X}
\newcommand*\Xd{\X_{\!\Delta}}
\newcommand*\Y{\mat Y}

@@

\usepackage{tikz}
\usetikzlibrary{shapes,snakes}
\usepackage{amsmath,amssymb}
% \usepackage{datatool}
% \usepackage{glossaries}

@@

%    \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
%    \setlength\marginparsep{3pt}
%    \setlength\marginparwidth{42pt}
%    \newlength\marginparwidthsmall
%    \setlength\marginparwidthsmall{\marginparwidth}
%    \addtolength\marginparwidthsmall{-7pt}
%    \newcommand\todo[1]{%
%       \addtocounter{todoidx}{1}%
%       {\color{Red}\fbox{\bf\thetodoidx{}}}%
%       \marginpar{%
%          {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
%          \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{#1}}}}
%

% correct bad hyphenation here
% \hyphenation{op-tical net-works semi-conduc-tor}

%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% GLOSSARY %
%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% \makeglossaries
% \input{../../Library/glossary}

%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% DOCUMENT START %
%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% \usepackage{yfonts}

\graphicspath{submission/final/gfx/}

\begin{document}

\title{An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}

\author{Jo~Inge~Buskenes,~\IEEEmembership{Student Member,~IEEE,} %
        Jon~Petter~\AA{}sen,~\IEEEmembership{Student Member,~IEEE,} %
        Carl-Inge~Colombo~Nilsen,~\IEEEmembership{Member,~IEEE,} %
        Andreas~Austeng,~\IEEEmembership{Member,~IEEE}%
\thanks{J. I. Buskenes, C.-I. C. Nilsen and A. Austeng are with the Department of Informatics, University of Oslo, Norway.}%
\thanks{J. P. \AA{}sen is with MI-Lab, Norwegian University of Science and Technology, Trondheim, Norway.}% <-this % stops a space
% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
}

% The paper headers
\markboth{IEEE Journal of Oceanic Engineering}%
{A Robust and Optimized GPU Implementation of the MVDR beamformer}

% Publishers ID mark:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
% Two limiting factors of modern phased array sonar imaging systems are the angular resolution and contrast. Adaptive beamformers target these limitatitons by weighting sensors dynamically to fit the impinging wavefield.

@@

% Keywords (normally not used for peer reviews)
\ifPeerReview\else
\begin{IEEEkeywords}
Adaptive beamforming, MVDR, Capon, sonar, active sonar, GPU, graphics processing unit, complexity.
\end{IEEEkeywords}
\fi}
% \fi

% make the title area
\maketitle

% This command fixes abstract positioning for compsoc articles:
\IEEEdisplaynotcompsoctitleabstractindextext

% (Optional) Add some extra info on cover page of peer review papers:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi

% Insert page break and insert second title (peer review mode)
\IEEEpeerreviewmaketitle

\maketitle

@@

\IEEEPARstart{D}{ata} driven methods have been introduced in various imaging fields over the years in attempts to improve image quality. One such method is the Minimum Variance Distortionless Response (MVDR) beamformer. When compared to conventional methods, MVDR is often able to improve image contrast and resolution, as shown in e.g. radar~\cite{Benitz1997}, ultrasound~\cite{Synnevag2007}, and active sonar~\cite{Blomberg2012a,Blomberg2011,Dursun2009,Lo2004}.

@@

Our implementation is adapted to a well sampled 100\;kHz wideband active sonar system operating in the near field. The system has a range resolution comparable to the sample period so our MVDR implementation runs in near single snapshot mode. To deal with this and to avoid signal cancellation we compute and average the covariance estimates from overlapping subarrays~\cite{Kailath1985}. This may resemble techniques such as "subarray MVDR"~\cite{Chapman1976}, but unlike that technique we are not using subarrays to reduce complexity but to make it feasible to use MVDR in an active system. In related work, we investigated using the same MVDR GPU implementation in cardiac ultrasound imaging~\cite{Asen2012,Asen2013}, and in active sonar imaging we have compared the performance of this implementation to comparable adaptive methods~\cite{Buskenes2013}.

@@

@@

@@

\ifPeerReview
\begin{figure}[H]\centering
\includegraphics[width=0.8\linewidth]{gfx/buske1\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/beamforming.eps}
\fi%
\caption{Beamforming principle. Signal signature is first removed by matched filtering. Then - before summation - a suitable set of delays, $\Delta$, and weights, $w$, are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{beamforming}
\end{figure}
Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (Fig. \ref{beamforming}). Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$. To simplify notation, we make the dependence on $\theta$ implicit from now on.

@@

Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield. In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}. This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{weights}
\end{gather}
where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\} \in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since we pre-steer our data to every pixel in the image we simplify (\ref{weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering. To estimate $\R$ we compute a sample covariance matrix $\eR$. In this computation we perform some degree of:
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985},
\item \emph{temporal averaging} over an interval comparable to the pulse length (1-5 samples) to maintain true speckle statistics~\cite{Synnevag2009a}, and
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
\end{itemize}%
%
% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009a}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}. 
%
Combined, these steps will also ensure a numerically well conditioned $\eR$.

We perform \emph{temporal} and \emph{spatial averaging} first and put the result in an intermediate sample covariance matrix $\breve{\R}$. To do this we need to segment our array into subarrays. If we let $x_l[n]$ represent the data vector from subarray $l$,
\begin{gather}
\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T,
\end{gather}
then $\breve{\R}$ can be calculated as:
\begin{gather}
\breve{\R}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n']\x_l\H[n'] \in\mathbb{C}^{L,L},\label{spatialR}
\end{gather}
where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.

@@

Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{z}). This weight set is applied to all the subarrays, before computing the beamformer output as in (\ref{z}). Or, equivalently, we may apply the weight set to the sum of all the subarrays:
\begin{align}
z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{finalZ}
\end{align}
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske2\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/implementation.eps}
\fi%
\caption{MVDR beamforming. For each pixel in range and azimuth,\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
3. and the beamformer output $z$ is computed from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
\end{figure}
As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
\begin{enumerate}
\item computing the sample covariance matrix $\eR$ in (\ref{finalR}),
\item computing $\eRi\1$ in (\ref{weights}), and
\item computing the beamformer output $z$ in (\ref{finalZ}).
\end{enumerate}
Next we will evaluate these steps in terms of arithmetic complexity, and then discuss their mappability to parallel hardware.

@@

If we neglect spatial and temporal averaging, then the computation of $\eR$ is reduced to a single outer product with complexity of O($M^2$), and the inversion is then of O($M^3$). This might lead us to believe that the inversion step is by far the most complex. But if we implement spatial and temporal averaging as in (\ref{spatialR}), then computing $\eR$ is of O($N_K N_L L^2$) and the inversion of O($L^3$). Computing $\eR$ is now the most complex operation whenever $N_LN_K>L$.
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske3\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_complexity.eps}
\fi%
\caption{Per-pixel computational complexity of the steps in MVDR beamforming (before any optimizations). To avoid signal cancellation in an active sonar system we usually set $L<\frac{M}{2}$, in which region the computation of $\eR$ dominates in terms of arithmetic complexity, especially when performing temporal averaging.}\label{mvdr_complexity}
\end{figure}
To visualize these relations, and include the effects of using complex numbers, we set up complexity formulas that account for the total number of arithmetic operations for each step in the MVDR process (see appendix \ref{mvdr_formulas}). The only excluded operation was partial pivoting used in the inversion step, but this should contribute marginally to the end result. The entire range of possible subarray sizes from $L\in[1,M]$ was finally evaluated, with temporal averaging set to $K\in\{0,1,2\}$, and the number of channels set to $M\in\{8,16,32\}$. The results are shown in \Fig{mvdr_complexity}.

Note how the computation of $\eR$ completely dominates at smaller subarray sizes, that solving $\eRi\1$ only plays a notable role for larger array and subarray sizes, and that the computation of $z$ has a negligible impact on computation time. Also notice how temporal averaging comes with a high computational penalty. This is because building $\eR$ is heavy on complex multiplications, which require 3 times as many arithmetic instructions as a complex addition, and a lot of these are repeated unnecessarily.

@@

@@
